Non-word repetition in Yélî Dnye

### 2021-08-20 manuscript.Rmd
first stab at R1

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before). In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. The study was designed to contribute to four broad research areas, three via direct results.

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills. Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

Second, we varied the length (in syllables) of non-words to contribute to growing research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer. To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]; and integrated those data with Yélî Dnye results from the present study in Figure \@ref(fig:fig-prevlit).

@@

```{r fig-prevlit, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years) and item length for comparable studies (2-4 indicating number of syllables,  2=dashed, 3=dotted, 4=dotted and dashed). Jaber-Awida (2018) reported on 20 Israeli Arabic learners (orange); Piazzalunga et al. (2019) reported on groups of 24-60 Italian learners (black); Stokes et al. (2006) on 15 Cantonese learners (blue); Vance et al. (2005) on 17-20 English learners (light green); Cristia et al. (2020) reported on groups of 4-6 Tsimane' learners (dark green); the present study reports on groups of 8-19 Yélî Dnye learners (purple). Central tendency is the mean except for Italian and Yélî Dnye (median); error is one standard error. Age has been slightly shifted for ease of inspection of different lengths at a given age."}

mycols = c( 
             "Arabic" = "orange",
             "Italian"   ="black",
             "Cantonese"   ="blue",
             "English"   ="green",
            "Tsimane'" = "darkgreen",
             "Yélî Dnye" = "purple")

mywd = c( 
             "Arabic" = 1,
             "Italian"   =1,
             "Cantonese"   =1,
             "English"   =1,
             "Tsimane'" = 2,
             "Yélî Dnye" = 4)

#just a trick to jitter age
jit_age=(pl_stacked$Age+(as.numeric(as.character(pl_stacked$Length))-2)/5)

plot(1,xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100),xlim=range(pl_stacked$Age))

for(thislang in levels(pl_stacked$Language)) for(thislength in levels(factor(pl_stacked$Length[pl_stacked$Language== thislang]))){
  myx=pl_stacked$Age[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myx=myx+(as.numeric(as.character(thislength))-2)/5 #jitter the length lines a little bit with respect to avg age
  points(pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]~myx,col=mycols[thislang],pch=as.character(thislength))
  myy=pl_stacked$Median[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  lines(myy~myx, col=mycols[thislang],lty=as.numeric(thislength),lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  myhb=myy+pl_stacked$se[pl_stacked$Language==thislang & pl_stacked$Length==thislength]
  for(i in 1:length(mylb)) {
    mylty=as.numeric(as.character(thislength))
    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang],lty=mylty) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang],lty=mylty) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang],lty=mylty) #higher error bar tip
    }
}

Our reading of this Figure is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.

@@

Fourth, these data contribute to the small literature using this task with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge]. There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores. In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations). Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.

@@

### Yélî Dnye phonology

@@

### The Yélî community

@@

### NWR design and analysis adaptations

@@

Throughout these analyses and in the Discussion, we will also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.

@@

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were first borns, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^3] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^2]

[^3]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.

@@

### NWR as a function of cross-linguistic phone frequency

@@

### Patterns in NWR mispronunciations

@@

### NWR scores as a function of item length

@@

### Factors structuring individual variation

@@

### Associations between NWR and cross-linguistic frequency

@@

### Item Length

@@

### Individual differences

@@

```{r fig-prevlit-overall, echo = FALSE, ig.pos = "T!", fig.width=8, fig.height=5.5, fig.cap = "NWR scores as a function of age (in years), averaged across multiple non-word lengths, as a function of children's native languages. The legend indicates language and the length of non-words (in syllables). Central tendency is mean; error is one standard error."}
mycols = c( 
             "Arabic" = "orange",
             "Cantonese"   ="blue",
             "English"   ="green",
             "Italian"   ="black",
             "Mandarin"   ="gray",
             "Persian"   ="pink",
             "Slovak"   ="lightblue",
            "Sotho"   ="magenta",
            "Spanish"   ="brown",
            "Swedish"   ="red",
            "Tsimane'" = "darkgreen",
             "Yélî" = "purple")

mywd = c( 
             "Arabic" = 1,
             "Cantonese"   =1,
             "English"   =1,
             "Italian"   =1,
             "Mandarin"   =1,
             "Persian"   =1,
             "Slovak"   =1,
            "Sotho"   =1,
            "Spanish"   =1,
            "Swedish"   =1,
             "Tsimane'" = 2,
             "Yélî" = 4)

plot(pl_overall$Overall~pl_overall$Age,col=mycols[pl_overall$Language],pch=1,
     xlab="Average age (years)",ylab="NWR scores (words)",ylim=c(0,100))

for(thislang in levels(factor(pl_overall$Language))) {
  myx=pl_overall$Age[pl_overall$Language==thislang ]
  myy=pl_overall$Overall[pl_overall$Language==thislang]
  lines(myy~myx, col=mycols[thislang],lwd=mywd[thislang]) #join points of same length & lang
  mylb=myy-pl_overall$se[pl_overall$Language==thislang]
  myhb=myy+pl_overall$se[pl_overall$Language==thislang]
  for(i in 1:length(mylb)) {

    lines(rep(myx[i],2),c(mylb[i],myhb[i]),col=mycols[thislang]) #error bar
    lines(c(myx[i]-.1,myx[i]+.1),c(mylb[i],mylb[i]),col=mycols[thislang]) #lower error bar tip
    lines(c(myx[i]-.1,myx[i]+.1),c(myhb[i],myhb[i]),col=mycols[thislang]) #higher error bar tip
    }
}

#add legend
#first half of langs
xmin=6
ymax=30
ymin=0
myy=seq(from=ymax,to=ymin,length.out=length(all_lang)/2)
names(myy)<-all_lang[1:(length(all_lang)/2)]
for(thislang in names(myy)){
  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
}

#second half
xmin=9.5
names(myy)<-all_lang[(length(all_lang)/2+1):length(all_lang)]
for(thislang in names(myy)){
  lines(c(xmin-.1,xmin+.1),c(myy[thislang],myy[thislang]),col=mycols[thislang]) #line sample
  text(xmin+.15,myy[thislang],paste(thislang,pl_overall$Lengths[pl_overall$Language==thislang][1]),col=mycols[thislang],pos=4)
} 
```

@@

>>>>

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with gender, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

TODO

- add subheadings (here, results, discussion)
- remove topics in intro that do not relate to aims of study

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood [e.g., @hazan2000development]. Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. This allows us to contribute data at the crossing of language typology, language acquisition, and individual variation, as presented in more detail below.

### What is NWR?

In a NWR task, participants hear a short word-like form that is phonologically legal but lexically meaningless in the language(s) they are learning. After hearing this non-word, the participant's task is to try to immediately and precisely repeat it. NWR has been used to seek answers to a variety of theoretical questions, including what the links  between phonology, working memory, and the lexicon are [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language;@chiat2015nonword;@armon2015assessing]. NWR scores are thought to reflect long-term phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

#### NWR design and analysis adaptations

<!-- ac2ac address R's comment here: NWR tasks and what they measure. The authors suggest that their task measures phonological development, I think this point needs to be sharpened and better justified. Yes, researchers' opinion on what the task measures vary greatly, and it is briefly over-viewed. Yet, the rational for what the novel task measures needs to be either more strongly justified, or down-toned. -->

#### NWR and typology

The first research area is at the crossing of typology and phonological development.  Previous work using NWR has preferred relatively universal and early-acquired phonemes [with exceptions including @gallagher2014identity], in part as a way to separate phoneme pronunciation from broader syllable structure and word-level prosodic effects [@gallon2007non] and in part because the test is sometimes used to measure working memory in the context of executive functions [@mulder2017early] rather than purely linguistic skills.

EXPLAIN MORE CLEARLY what is the intuition here

Here, we investigate repetition of non-word items containing cross-linguistically common and cross-linguistically rare phonetic targets. Specifically, we included a subset of non-word items with typologically rare sounds to ask whether these sounds are disadvantaged in the perception-production loop involved in NWR.

<!-- ac2ac address R's comment: (2) Language typology (universal and early-acquired phonemes / cross-linguistically common and cross-linguistically rare phonetic targets). Here I would recommend consulting a Chapter by Shula Chiat on language-universal and vs. language-specific NWR tasks. -->
<!-- Chiat, S. (2015). Non-Word Repetition. In Armon-Lotem, S., de Jong, J., & Meir, N. (Eds.). Methods for assessing multilingual children: disentangling bilingualism from Specific Language Impairment (pp.125-150). Bristol: Multilingual Matters. -->

#### Length effects on NWR

@@

The third research area we contribute data to is research looking at the impact of word length on NWR repetition, and what this may reflect about phonological development within specific languages. Some work documents much lower NWR scores for longer, compared to shorter, items [e.g., among Cantonese-learning children; @stokes2006nonword], whereas differences are negligible in other studies  [e.g., among Italian learners; @piazzalunga2019articulatory]. It is possible that differences are due to language characteristics, including the modal length of words in the language and/or in child-directed speech in that culture. In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words are modally longer.

To see whether there were broad generalizations that could be drawn from previous literature fitting these predictions, we inspected NWR papers in a variety of languages which reported NWR scores separately for different word lengths, as described in Supplementary Materials [@cristia2021supplementary; SM1]. We found data for learners of Israeli Arabic  [@jabere2018xperiment]; Cantonese  [@stokes2006nonword]; English  [@vance2005speech]; Italian  [@piazzalunga2019articulatory]; and Tsimane'  [@cristia2020infant]. 
Our reading of that work is that, although there is cross-linguistic (or cross-sample) variation in length effects, these do not systematically line up with expected word length in different languages. For instance, the difference in NWR scores for 2- versus 3-syllable items (averaging across age groups) is largest in Tsimane' (~`r round(difs[difs$Group.1=="Tsimane'","x"])`%) and Arabic (~`r round(difs[difs$Group.1=="Arabic","x"])`%), which tend to have longer words, as does Italian, where the difference between 2- and 3-syllable items was only ~`r round(difs[difs$Group.1=="Italian","x"])`%. Similarly, two languages that are often described as heavily biased towards monosyllables show diverse length effects (Cantonese ~`r round(difs[difs$Group.1=="Cantonese","x"])`% versus English ~`r round(difs[difs$Group.1=="English","x"])`%). Given the paucity of research looking at this question, and the diversity of current results, we do not approach this issue within a hypothesis-testing framework but sought instead to provide one more piece of data on the question, which may be re-used in future meta- or mega-analytic analyses.

#### Individual variation effects on NWR

@@

#### NWR as a function of language and culture

The fourth research goal we pursued is to use NWR with non-Western, non-urban populations, speaking a language with a moderate to large phonological inventory [see @maddieson2005correlating for a broad classification of languages based on inventory size]. Indeed, NWR has seldom been used outside of urban settings in Europe and North America [with exceptions including @gallagher2014identity; @cristia2020infant], nor with languages having large phonological inventories [e.g., more than 34 consonants and 7 vowel qualities @maddieson2013vowel; @maddieson2013consonant; with no exceptions to our knowledge].

There are no theoretical reasons to presume that the technique will not generalize to these new conditions. That said, @cristia2020infant recently reported relatively lower NWR scores among the Tsimane', a non-Western rural population, interpreting these findings as consistent with the hypothesis that lower levels of infant-directed speech and/or low prevalence of literacy in a population could lead to population-level differences in NWR scores.

In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading, better than rhyme awareness for instance [e.g., @gathercole1991differentiating]. Therefore, it may not be a general index of phonological development, but reflect only certain non-universal skills. Indeed, @cristia2020infant present the task as being a good index of the development of "short-hand-like" representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations.

Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of language typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).

Regarding potential language differences, we note that the very fact that studies compose items by varying syllable structure and word length, while preferring relatively simple and universal phones (notably relying on point vowels, simple plosives, and fricatives that are prevalent across languages, like /s/) may indicate a bias towards Indo-European languages, where syllable structure and word length are indeed important structural dimensions. This bias is, of course, implicit and unintentional, arising as researchers working in other languages attempt to build items that conform to the descriptions of the first investigations using the method, who tend to involve English participants.

### Yélî Dnye phonology and community

@@

#### Yélî Dnye phonology

@@

#### The Yélî community

@@

@@

Throughout these analyses and in the Discussion, we also have in mind our fourth goal, namely integrating NWR results across samples varying in language and culture.

@@

We tested a total of `r dim(demo)[1]` children from `r length(levels(factor(demo$familyID)))` families spread across four hamlet regions. We excluded test sessions from analysis for the following reasons: refused participation or failure to repeat items presented over headphones even after coaching (N=`r sum(demo$included=="didn\xeat repeat")`), spoke too softly to allow offline coding (N=`r sum(demo$included=="spoke too softly")`), or were 13 years old or older (N=`r sum(demo$Age >12 | demo$included=="too old",na.rm=T)`; we tested these teenagers to put younger children at ease). The remaining `r dim(inc)[1]` children (`r sum(inc$Sex=="G")` girls) were aged from `r min(inc$Age, na.rm=T)` to `r max(inc$Age, na.rm=T)` years (M = `r round(mean(inc$Age, na.rm=T),1)` years, SD = `r round(sd(inc$Age, na.rm=T),1)` years).  In terms of birth order, `r birthorder.tab[[1]]` were born first, `r birthorder.tab[[2]]` second,  `r birthorder.tab[[3]]` third,  `r birthorder.tab[[4]]` fourth,  `r birthorder.tab[[5]]` fifth, and `r birthorder.tab[[6]]` sixth, with birth order missing for `r sum(is.na(inc$birthOrder))` children. These children were tested in a hamlet far from our research base, and we unfortunately did not ask about birth order before leaving the site. Maternal years of education averaged `r round(mean(inc$matEd,na.rm=T),2)` years (range `r round(min(inc$matEd,na.rm=T),2)`-`r round(max(inc$matEd,na.rm=T),2)` years).[^2] We also note that there were `r sum(inc$monolingual=="yes", na.rm=T)` children only exposed to Yélî Dnye at home and `r sum(inc$monolingual=="no", na.rm=T)` children exposed to Yélî Dnye plus one or more other languages at home.[^3]

[^2]: We asked for mothers' highest completed level of education. We then record the number of years entailed by having completed that level under ideal conditions.

@@

@@

### NWR and typology: NWR as a function of cross-linguistic phone frequency

@@

#### Follow-up analyses: Patterns in NWR mispronunciations

@@

### Length effects on NWR

@@

### Individual variation effects on NWR

@@

### NWR and typology

We looked at two things: Associations between NWR and cross-linguistic frequency, and qualitative mispronunciation patterns. RESTRUCTURE TO MAKE THIS CLEARER!

@@

###  Length effects and NWR

@@

### Individual variation effects on NWR

@@

FIGURE REMOVED

@@

## Data, code and materials availability statement

All data, code, and materials are available from [https://osf.io/5qspb/](https://osf.io/5qspb/)
